Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-20 paper/ms.tex
Added discussion of kernel approximation.

\include{vc}
\usepackage{microtype}
\usepackage{amsmath}

@@

The kernel is modeled as a sum of damped harmonics,
\begin{equation}
K(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
and zero otherwise.

@@

Consequently, we use the fact that $\exp{i\beta_I \tau} =  \cos{\beta_I \tau} + i \sin{\beta_I \tau}$ to
rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
may be replaced by a $2\times 2$ matrix, $ \begin{pmatrix} 0 & 1 \\ -1 & 0\end{pmatrix}$, so that
$\cos{\beta_I \tau} + i \sin{\beta_I \tau}$ may be rewritten as the rotation matrix
$\begin{pmatrix} \cos{\beta_I \tau} & \sin{\beta_I \tau} \\ -\sin{\beta_I \tau} & \cos{\beta_I \tau}\end{pmatrix}$.
We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex 
equations, resulting in a real extended matrix that is the same size as the complex matrix.
This transformation leads to about a factor of two in computational savings, but has the
advantage of allowing the automatic differentiation due to the use of real arithmetic.

\subsection{Kernel components}

>>>>

\include{vc}
\usepackage{microtype}

@@

The kernel is modeled as a sum of damped harmonics,
\begin{equation}
K(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
of the measured times if the correlated components are zero.

@@

Consequently, we use the fact that $\exp{(i\beta_I \tau)} =  \cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ to
rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
may be replaced by a $2\times 2$ anti-symmetric matrix,
\begin{equation}
\left(\begin{tabular}{ll}
0 & 1 \\
-1 & 0
\end{tabular}\right),
\end{equation}
while real numbers use the $2\times 2$ identity matrix, $\mathbf{I}$, 
so that $\cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ may be rewritten as the rotation matrix
\begin{equation}
\left(\begin{tabular}{ll} 
$\cos{(\beta_I \tau)}$ & $\sin{(\beta_I \tau)}$ \\ 
$-\sin{(\beta_I \tau)}$ & $\cos{(\beta_I \tau)}$ 
\end{tabular}\right).
\end{equation}
The algebraic properties of these matrices are identical to the complex numbers, but only involve
real components.
We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex 
equations, resulting in a real extended matrix that is the same size as the complex matrix.
This transformation leads to about a factor of two in computational savings, but has the
additional advantage of allowing the automatic differentiation due to the use of real arithmetic.

\section{Kernel components}

A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
plus the white noise component:
\begin{equation}
K(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
\end{equation}
Note that this encompasses all smooth, time-staionary kernels, and generally $c_1 \le 0$ as
correlated noise initially decreases with separation in time.  We wish to approximate this
as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
can carry out the approximation in terms of sums of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)}$,
which has a Taylor series expansion:
\begin{equation}
\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} = 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 -
\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
\end{equation}
which may be continued to any order in $\tau$ as necessary.
If an approximation to $K(\tau)$ is desired in a Taylor series to some high order $N$, this
requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
chosen to match the coefficients $c_j$.  Not that for $N=1$ we only require a damped exponential
kernel with $\beta_I = 0$.

Another way to view this is in the power spectrum of the Gaussian Process, which is the
Fourier Transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
power spectrum is: {\color{red} Check that this is true.}
\begin{equation}
P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
\end{equation}
where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
that a smooth distribution may be approximated arbitrarily precisely by summing together
Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
a peak in the power spectrum is narrow at a specific frequency, then a small value of 
$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
can be arranged to cancel in the wings, giving features in the power spectrum that are
narrower in frequency.

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item Constant kernel
\item The squared exponential kernel has the behavior of declining exponentially at large
separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
approximate a Gaussian kernel.  For example, to 
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0)
= (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This may be
well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are too large.  One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be an excellent description of stellar granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.
\item Matern 5/2 kernel.
\item Cosine kernel
\item exp-sin-sq kernel
\end{itemize}
